---
title: "Soil Mechanics"
author: "Maximiliano Garnier Villarreal"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
vignette: >
  %\VignetteIndexEntry{Soil Mechanics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(GMisc)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  fig.align = "center",
  out.width = "80%"
)
```

This document highlights some functions of the GMisc package related to soil mechanics problems.

# Bearing capacity

The idea here is to provide a function that solves the bearing capacity equation for different types of footings, giving as a result the allowable bearing capacity ($q_a$):

$$
q_a = \frac{0.5 \ B \ \gamma \ N_{\gamma} \ s_{\gamma} + D \ \gamma \ N_q + c \ N_c \ s_c}{FS}
$$

where $B$ is the width in meters, $D$ is the embedment depth in meters, $FS$ is the desired factor of safety, $c$ is the cohesion in kPa (referred as `tau0` in the function), $\gamma$ is the unit weight in $kN/m^3$, and the rest of the parameters are correction factors dependent on these input values, the internal friction angle $\phi$, the type of footing ("strip", "square", "rectangular","circular"), and the depth to the water level.

The function that solves the above equation is `bearing_capacity`, with the mentioned arguments but specifying the wet unit weight, saturated unit weight for their respective use. The function outputs a table and graph for the results.

The `B` and `D` parameters can be vectors for multiple cases comparisons or single values for a single case estimate. If `FS = 1` then the allowable and ultimate bearing capacities are the same ($qa = qu$).

Different scenarios can be analyzed:

* For a total stress analysis (TSA) in a cohesive soil (plastic silts and clays) set the friction angle equal to zero (`phi = 0`) and the cohesion equal to the undrained shear strength (`tau0 = Su`).
* For an effective stress analysis (ESA) in a cohesive soil use the effective friction angle ($\phi'$) and effective cohesion ($c'$).
* For a coarse-grained soil (gravels, sands, and non-plastic silts) usually TSA = ESA, and the friction angle and cohesion should be used, and if the material has no-cohesion then set cohesion equal to zero (`tau0 = 0`)

We can define a set of parameters to use in different examples as follows:

```{r}
B = seq(0.5, 2, 0.25)
D = seq(0, 2, 0.25)
L = NULL
gamma.h = 15.5
gamma.s = 18.5
tau0 = 10
phi = 30
FS = 3
wl = 1
```

The first example uses the default "strip" footing:

```{r}
bearing_capacity(B, D, L, gamma.h, gamma.s, tau0, phi, wl, FS)
```

The second example uses the a "square" footing:

```{r}
bearing_capacity(B, D, L, gamma.h, gamma.s, tau0, phi, wl, FS, footing = "square")
```

And the last example uses a "rectangular" footing, where is needed to define the `L` argument (the length of the footing in meters):

```{r}
bearing_capacity(B, D, L = 3, gamma.h, gamma.s, tau0, phi, wl, FS, footing = "rectangular")
```

# Induced stress

The function `induced_stress` calculates and plots the stress distribution under the center of a footing of different shapes ("strip", "square", "rectangular", "circular"), using the solutions from Boussinesq. The arguments of the function are `qs`: the stress applied by the footing, `B`: the width of the footing in meters, `L`: the length of the footing in meters (applies only for rectangular footings), and `z.end`: the depth down to which calculate the stress. The function returns two plots: one static (`$Plot`) and one interactive (`$Plotly`), the last one is the one shown here.

The first example shows the plot for the default "strip" footing, and one value of width:

```{r warning=FALSE}
qs = 1
B = 1
L = NULL
z.end = 8
induced_stress(qs, B, L, z.end)$Plotly
```

The second example plots the results for a "square" footing and different widths:

```{r}
qs = 1
B = c(1, 2, 4)
L = NULL
z.end = 8
induced_stress(qs, B, L, z.end, footing = "square")$Plotly
```

The last example plots the results for a "rectangular" footing and different lengths:

```{r}
qs = 1
B = 1
L = c(1, 2, 4)
z.end = 8
induced_stress(qs, B, L, z.end, footing = "rectangular")$Plotly
```

# Direct shear

The functions for direct shear are more for visualizing the results rather than to process the. The user should analyze the raw data and provide the normal stress (`sig.n`) - shear stress pairs (`tau`).

The first function plots the resulting $(\sigma,\tau)$ pairs and adds a line of best fit for the Mohr-Coulomb criterion ($\tau=c+\sigma_n tan\phi$). If the line of best fit estimates an intercept of less than 0 ($c<0$), the function automatically assigns $c=0$, as this is not a possible solution. Also, the estimated parameters ($c,\phi$) are added to the plot.

```{r warning=FALSE}
DS_plot(sig.n = c(80,237,395), tau = c(127,345,475))
```

The second function adds on top of the first one by drawing the Mohr circles for each of the $(\sigma,\tau)$ pairs.

```{r warning=FALSE, message=FALSE}
DS_Mohr_plot(sig.n = c(80,237,395), tau = c(127,345,475))
```

# Mohr circle

The idea here is to provide a way to solve simple Mohr circle's problems for the given normal stresses, shear stress, and rotation angle. The function returns a plot and the results for the given problem and parameters.

We can define some values to use:

```{r}
sigx = 143.6
sigy = 100.5
tauxy = -14.4
theta = -35
```

where `sigx` corresponds to the mayor normal stress, `sigy` corresponds to the minor normal stress, `tauxy` corresponds to the shear stress, and `theta` is the rotation angle for which to find the stresses.

In the first example the rotation is not applied:

```{r}
Mohr_Circle(sigx, sigy, tauxy)
```

In the second example the rotation is applied:

```{r}
Mohr_Circle(sigx, sigy, tauxy, theta)
```

# Shinyapp

A shinyapp using these functions can be accessed at https://maximiliano-01.shinyapps.io/soil_mechanics/.
